A general and efficient representation of Ancestral Recombination Graphs

### 2022-09-13 paper.tex
Comments with paragraph summaries
Merge eARG and gARG sections

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process, the classical definition of an ARG data structure is
in terms of \emph{events}. Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In order to distinguish this form of ARG from what we will be discussing
in later sections we refer to it as an ``event ARG``, or eARG for brevity.
An example of an eARG
is given in Fig.~\ref{fig-arg-data-structure}B.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage (e.g., node \noderef{e}
in Fig.~\ref{fig-arg-data-structure}), and in a recombination
event a lineage is split into two independent
ancestral lineages (e.g., node \noderef{d}
in Fig.~\ref{fig-arg-data-structure}). The final vital
detail---which distinguishes an arbitrary graph with the required topological
properties from an eARG---is that we associate a breakpoint with each recombination
event. This information is sufficient to uniquely
define the local genealogical trees at every position along the genome,
which is the basic requirement for a data structure encoding
genome-wide genetic ancestry.

@@

\node [anchor=north west] at (-1.5,6) {B};
\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};

@@

\node [anchor=north west] at (-14,6) {A};
\node [nodelabel,anchor=north west] at ($(-13,5)$) {
\begin{tabular}{c|c}
% \multicolumn{2}{c}{Breakpoints}\\
Nodes & Breakpoints\\
\hline
$\textsf{a}$ & $\varnothing$ \\
$\textsf{b}$ & $\varnothing$ \\
$\textsf{c}$ & $\varnothing$ \\
$\textsf{d}$ & $x$ \\
$\textsf{e}$ & $\varnothing$ \\
$\textsf{f}$ & $\varnothing$ \\
$\textsf{g}$ & $\varnothing$ \\
\end{tabular}};

\node [nodelabel,anchor=north west] at ($(-7,5)$) {
\begin{tabular}{l}
Edges\\
\hline
$(\textsf{a}, \textsf{e})$ \\
$(\textsf{b}, \textsf{d})$ \\
$(\textsf{c}, \textsf{f})$ \\
$(\textsf{d}, \textsf{e})$ \\
$(\textsf{d}, \textsf{f})$ \\
$(\textsf{e}, \textsf{g})$ \\
$(\textsf{f}, \textsf{g})$ \\
\end{tabular}};

\node [anchor=north west] at (9,6) {C};
\node [nodelabel,anchor=north west] at ($(10,5)$) {
\begin{tabular}{ll}
Edges & Intervals\\
\hline
$(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
$(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
$(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
$(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
$(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
$(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
$(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
\end{tabular}};

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A classical ARG (subfigure B), contrasting node with edge breakpoint annotations.
(A) recombination breakpoint information encoded on \emph{nodes}.
This encoding is limited in the patterns of inheritance that can be
described and requires that edges be ordered to avoid ambiguity.
(C) breakpoint annotations associated with \emph{edges}.
Associating the interval of genome inherited by the child node
from the parent with the corresponding edge allows
any pattern genetic inheritance to be described and does
not require edges to be ordered. In this example, each edge is associated
with a single interval, but more generally, multiple non-overlapping intervals
can be associated with a single edge.
}
\end{figure}

% What do we mean by an eARG, formally?
To be concrete, we can formally define the
classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
is an ordered list of edges defining a directed acyclic graph and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}$
is a function mapping nodes to recombination breakpoints,
such that $\sigma(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
$\sigma(u) = \varnothing$ otherwise.
(This is equivalent to having different ``types''
of node, where breakpoints are associated with recombination
nodes only).
Each edge $e_j = (c, p)$ describes an edge between
child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
% For each recombination node $u$ we must
% have exactly two edges in $e$ such that $u$ is the child.
This encoding of an eARG is equivalent to the output
of programs such as
\texttt{ARGweaver}~\citep{rasmussen2014genome}
and \texttt{KwARG}~\citep{ignatieva2021kwarg},
and an example is given in Fig.~\ref{fig-arg-data-structure}A.
Note that the description above captures only the
graph topology: if we also wish to know the branch lengths we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with nodes:
we need to know from which parent the node inherited
the segments of genome to the left and right of the breakpoint.
The simplest way to break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG.
To recover the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
when using this representation and simulating an ARG backwards in time,
the first event older than a recombination may involve the lineage carrying the
ancestry to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course, but depending on the ordering of
otherwise indistinguishable objects is generally problematic. In practice,
several methods explicitly associate information with the outbound edges
of a recombination event
to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.

@@

The eARG encoding consists of two types of event
(common ancestor and recombination) but nature is not so parsimonious
and there are many other processes that will affect the
ancestry of samples.
For example, gene conversion between homologous
chromosomes~\citep{wiuf2000coalescent,chen2007gene} is an important
evolutionary force in many species~\cite[e.g.][]{yang2012great}.
% Some refs for where people are inferring ancestry in the
% presence of GC? Isn't this what ClonalFrame etc are all about?
Similarly, the eARG encoding cannot easily represent multiple
recombinations on a chromosome, or the transmission of multiple
chromosomes.
What we can infer about ancestral histories and
processes should not be limited by the data structure that we
use to represent such ancestry. The eARG is simply one way
to \emph{encode} such history, and, as we have shown,
one that has significant limitations of expressivity
derived from its historical origins in the idealised
coalescent with recombination model.

@@

\section*{Genome ARGs}
As we have seen in the previous section, the focus
on a limited palette of events makes describing currently-available
datasets and known patterns of genetic inheritance as an eARG
cumbersome at best.
There is no fundamental reason that the patterns of genetic inheritance
that we wish to describe via an ARG must be in terms
of discrete \emph{events}, however. Recent discussions have illustrated the
relationship between ARGs and
the pedigree~\citep{mathieson2020ancestry,brandt2021evaluation},
and the deep connections between the structures are
well understood
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}.
Following \cite{mathieson2020ancestry},
rather defining the inheritance
of genetic material in terms of \emph{events} as we do in an eARG,
we can instead define an ARG data structure in terms of
\emph{genomes} and intervals of genetic inheritance
along the edges.
To distinguish from the classical Griffiths eARG,
we refer to this encoding of an ARG data structure
as a ``Genome ARG'', or gARG.
The distinction between nodes representing events
and genomes may seem pedantic at first, but as we
see in later sections, we gain a great deal of flexibility
and computational efficiency by decoupling our representation
of genetic ancestry from the events that generated it.

@@

A gARG is a graph in which nodes represent monoploid genomes (i.e.,
haploids have one genome, diploids two, etc.) and edges represent
genetic inheritance between an ancestor and a descendant.
Each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance).
Fig.~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
showing how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
A diploid pedigree is shown here for concreteness, but
there is no limitation on mating system, ploidy, or age structure
in the gARG encoding.
In the diploid setting, a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty.

Formally, we can define a gARG as a set of edges $E$, where each
element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section).
Deriving the local tree at a point $x$
follows largely the same pattern as for an eARG but is somewhat
simpler because coordinate information is directly associated with
edges. Starting with the sample nodes $S$ we traverse
rootwards through the graph from child to parent. At each node $u$, we find an
edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
and move to the next node $p$.

% This paragraph is a bit weak and muddled, needs a few more passes.
% In particular, the "Finally" bit doesn't really seem to follow
% from the others. Is there a more direct way we can address these points ?
There are some interesting and important consequences to this focus on genomes
rather than events.
Firstly, a gARG may contain nodes that are both samples \emph{and}
internal nodes in the local trees.
This ability is useful when we have pedigree information
along with genetic data
\cite[e.g.][]{hayes20191000,RosFreixedes2022,anderson2022genes},
where we may have many generations of internal sample nodes
whose genomes have been sequenced.
Another situation in which it is important that
we do not assume samples are all contemporary ``leaf'' nodes
(or that all internal nodes are either coalescences or recombinations)
is when we are incorporating ancient genomes into ARG
inference~\citep{speidel2021inferring,wohns2022unified}.
Similarly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{h} is the direct ancestor of \textsf{d} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{d} descends directly from \textsf{k} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective.
Secondly, because we do not classify nodes by event type
(common ancestor or recombination) there is no limitation
on the complexity that can be modelled---multiple recombinations
and coalescences can happen simultaneously in the same genome,
a common occurance in large deeply sequenced
pedigrees~\citep{hayes20191000,RosFreixedes2022}.
Finally, there is a subtle and important point about how recombination
is represented: which genome in the organismal lifecycle
do nodes represent?
In principle, we can interpret nodes as representing genomes
at any point in the life cycle [supplementary Figure ?]. Hudson
views nodes as representing gametes~\citep{hudson1983properties}.
However, we argue that the most intuitive
approach is to let nodes represent the genomes present in diploid
individuals in each generation, as in
Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
is not represented by a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place
(e.g.\ nodes \textsf{e} and \textsf{f}), and a
single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would be associated with a
recombination node in an eARG. [TODO finish up here with more
clarity on what this means in terms of the required specifity
of recombination events, and put in a forward ref to the simplification
section where we talk about stacking up recombination events
and their identifiability.]

For simplicity, throughout the rest of the paper we will contrast the
classical eARG (as defined in the XXX section) with the gARG definition
here. However, it is important to note that many of the benefits of the
gARG encoding that we discuss are due to the encoding of inheritance via
edges rather than nodes.
Whether one regards nodes as genomes or events is orthogonal,
and something of a philosophical question which makes little
difference from a computational perspective.

\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
The edges in a gARG represent a path of genetic inheritance from
parent to child (and ultimately from cell to cell) through some
number of generations. The edges are ``annotated''
with the genome coordinates over which inheritance occurs. There are
two ways in which we can interpret these annotations, and the
distinction is subtle and important.
The first interpretation is to regard the annotations as describing
the complete set of genomic intervals inherited by the descendant node from the
ancestor node. The second is to view the edge annotations as representing
the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample that is propagated through
the graph from descendant to ancestor.
These alternatives are illustrated in Fig~\ref{fig-ancestry-resolution}.

>>>>

\section*{Event vs Genome ARGs}\label{eARG}
In keeping with its original derivation in terms of a stochastic
process, the classical definition of an ARG data structure is
in terms of events.
Nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage, and in a recombination
event a lineage is split into two independent
ancestral lineages.
% The final vital
% detail is that we associate a breakpoint with each recombination
% event.
% This information is sufficient to uniquely
% define the local genealogical trees at every position along the genome,
% which is the basic requirement for a data structure encoding
% genome-wide genetic ancestry.
As illustrated in Fig.~\ref{fig-arg-data-structure}, this information
is usually concretely encoded by associating a recombination breakpoint
(which is null for common ancestor events) and an ordered list of parents
with each node. We refer to such a  data structure as an
``event ARG'' (eARG).
To recover the local tree at genome coordinate $x$ (the most
fundamental operation required for an ARG data structure),
we traverse the graph rootwards from the leaves.
At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
The order in which parents are listed at a recombination node is
therefore significant, telling us
from which parent the segments of genome to the left and right of the breakpoint
were inherited.

% This ordering requirement, while straightforward
% to describe, has some practical drawbacks. For example,
% % Surely the meaning is clear from the context here and we don't need
% % qualify the meaning of ``ARG''
% when using this representation and simulating an ARG backwards in time,
% the first event older than a recombination may involve the lineage carrying the
% ancestry to the right of the breakpoint rather
% than the left, and so we cannot emit edges as they are generated.
% Such issues can be worked around, of course, but depending on the ordering of
% otherwise indistinguishable objects is generally problematic.

@@

% \node [anchor=north west] at (0,6) {A};
\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x$};
\foreach \u/\lab in {s0/$\textsf{a}$, s1/$\textsf{b}$, s2/$\textsf{c}$} \node[nodelabel,anchor=north] at (\u) {\lab};
\foreach \u/\lab in {s4/$\textsf{e}$} \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {s5/$\textsf{f}$} \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/$\textsf{d}$, s6/$\textsf{g}$} \node[nodelabel,anchor=south] at (\u) {\lab};

@@

% \node [anchor=north west] at (9,6) {B};
\node [nodelabel,anchor=north west] at ($(10,5)$) {
\begin{tabular}{c|c|l}
% \multicolumn{2}{c}{Breakpoints}\\
Node & Breakpoint & Parents\\
\hline
$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{g}$ & $\varnothing$ & []\\
\end{tabular}};

% \node [anchor=north west] at (9,6) {C};
% \node [nodelabel,anchor=north west] at ($(10,5)$) {
% \begin{tabular}{ll}
% Edges & Intervals\\
% \hline
% $(\textsf{a}, \textsf{e})$ & $(0, L]$ \\
% $(\textsf{b}, \textsf{d})$ & $(0, L]$ \\
% $(\textsf{c}, \textsf{f})$ & $(0, L]$ \\
% $(\textsf{d}, \textsf{e})$ & $(0, x]$ \\
% $(\textsf{d}, \textsf{f})$ & $(x, L]$ \\
% $(\textsf{e}, \textsf{g})$ & $(0, L]$ \\
% $(\textsf{f}, \textsf{g})$ & $(0, L]$ \\
% \end{tabular}};

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A graphical depiction of a classical event ARG and a concrete
encoding of the information. Node \noderef{d} is a recombination
event as it is associated with a non-null breakpoint, and has
two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
common ancestor events.
}
\end{figure}

% % What do we mean by an eARG, formally?
% To be concrete, we can formally define the
% classical Griffiths eARG as a tuple $(e, \sigma)$, where $e$
% is an ordered list of edges defining a directed acyclic graph and
% $\sigma: \mathbb{N} \rightarrow \mathbb{N}$
% is a function mapping nodes to recombination breakpoints,
% such that $\sigma(u) = x$
% if $u$ is a recombination event with breakpoint $x$ and
% $\sigma(u) = \varnothing$ otherwise.
% (This is equivalent to having different ``types''
% of node, where breakpoints are associated with recombination
% nodes only).
% Each edge $e_j = (c, p)$ describes an edge between
% child node $c$ and parent $p$, where $c, p \in \mathbb{N}$.
% % For each recombination node $u$ we must
% % have exactly two edges in $e$ such that $u$ is the child.
% This encoding of an eARG is equivalent to the output
% of programs such as
% \texttt{ARGweaver}~\citep{rasmussen2014genome}
% and \texttt{KwARG}~\citep{ignatieva2021kwarg},
% and an example is given in Fig.~\ref{fig-arg-data-structure}A.
% Note that the description above captures only the
% graph topology: if we also wish to know the branch lengths we need
% an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
% which defines the time of each node.

% In practice,
% several methods explicitly associate information with the outbound edges
% of a recombination event
% to resolve the issue~\citep{lyngso2005minimum,ignatieva2021kwarg}.

@@

% JK: Removing stuff about the limitations of the encoding here
% because it's reasonably straightforward to see an extension in
% which the number of breakpoints is in less than the number of
% parents, and it starts muddying the waters a bit.

% The eARG encoding consists of two types of event
% (common ancestor and recombination) but nature is not so parsimonious
% and there are many other processes that will affect the
% ancestry of samples.
% For example, gene conversion between homologous
% chromosomes~\citep{wiuf2000coalescent,chen2007gene} is an important
% evolutionary force in many species~\cite[e.g.][]{yang2012great}.
% % Some refs for where people are inferring ancestry in the
% % presence of GC? Isn't this what ClonalFrame etc are all about?
% Similarly, the eARG encoding cannot easily represent multiple
% recombinations on a chromosome, or the transmission of multiple
% chromosomes.
% What we can infer about ancestral histories and
% processes should not be limited by the data structure that we
% use to represent such ancestry. The eARG is simply one way
% to \emph{encode} such history, and, as we have shown,
% one that has significant limitations of expressivity
% derived from its historical origins in the idealised
% coalescent with recombination model.

@@

% \section*{Genome ARGs}
There is no fundamental reason, however, that the data structure we use
to describe patterns of genetic inheritance must be in terms
of specific causal events that generated those patterns.
Let us define a genome as the complete set of genetic material that a child
inherits from one parent (i.e., haploids have one genome, diploids two, etc.).
Then, a ``genome ARG'' (gARG) is a graph in which nodes represent
monoploid genomes  and edges represent
genetic inheritance between an ancestor and a descendant,
where each edge is annotated with the coordinates
of the genomic intervals over which genetic inheritance occurs
(the next section explores the subtleties of what we mean by
genetic inheritance). To recover the local tree for a genomic
position $x$, we traverse the graph rootwards from the leaves
as in an eARG. At a particular node, we then follow the outbound
edge associated with the genomic interval containing $x$.

% Recent discussions have illustrated the
% relationship between ARGs and
% pedigrees~\citep{mathieson2020ancestry,brandt2021evaluation},
% and the deep connections between the structures are
% well understood
% \citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
% speed2015naturereviewsgenetics}.
Fig.~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG,
showing how the genomes embedded in an example (highly inbred)
pedigree correspond to the nodes in a gARG, along with the corresponding
local trees.
A diploid pedigree is shown here for concreteness, but
there is no limitation on mating system, ploidy, or age structure
in the gARG encoding.
In the diploid setting, a genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.
Arbitrary patterns of genetic inheritance (gene conversions, multiple recombinations,
etc) can be expressed without difficulty.

% To distinguish from the classical Griffiths eARG,
% we refer to this encoding of an ARG data structure
% as a ``Genome ARG'', or gARG.
% The distinction between nodes representing events
% and genomes may seem pedantic at first, but as we
% see in later sections, we gain a great deal of flexibility
% and computational efficiency by decoupling our representation
% of genetic ancestry from the events that generated it.

@@

% Formally, we can define a gARG as a set of edges $E$, where each
% element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
% $c$ is the child node, $p$ is the parent node and $I$ is the set of
% disjoint genomic intervals $(\ell, r]$ over which genome $c$ inherits from $p$ (we may
% also phrase this in terms of the ancestral material of a sample---see
% the next section).
% Deriving the local tree at a point $x$
% follows largely the same pattern as for an eARG but is somewhat
% simpler because coordinate information is directly associated with
% edges. Starting with the sample nodes $S$ we traverse
% rootwards through the graph from child to parent. At each node $u$, we find an
% edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$
% and move to the next node $p$.

% This paragraph is a bit weak and muddled, needs a few more passes.
% In particular, the "Finally" bit doesn't really seem to follow
% from the others. Is there a more direct way we can address these points ?
% There are some interesting and important consequences to this focus on genomes
% rather than events.

% Firstly, a gARG may contain nodes that are both samples \emph{and}
% internal nodes in the local trees.
% This ability is useful when we have pedigree information
% along with genetic data
% \cite[e.g.][]{hayes20191000,RosFreixedes2022,anderson2022genes},
% where we may have many generations of internal sample nodes
% whose genomes have been sequenced.
% Another situation in which it is important that
% we do not assume samples are all contemporary ``leaf'' nodes
% (or that all internal nodes are either coalescences or recombinations)
% is when we are incorporating ancient genomes into ARG
% inference~\citep{speidel2021inferring,wohns2022unified}.
% Similarly, a gARG may contain nodes that
% do not correspond to any particular event in the ancestry of the samples.
% For example, node \textsf{h} is the direct ancestor of \textsf{d} in
% Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
% coalescence or a recombination. Such nodes would usually be removed
% from the graph so that \textsf{d} descends directly from \textsf{k} (see
% the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
% a representational perspective.
% Secondly, because we do not classify nodes by event type
% (common ancestor or recombination) there is no limitation
% on the complexity that can be modelled---multiple recombinations
% and coalescences can happen simultaneously in the same genome,
% a common occurance in large deeply sequenced
% pedigrees~\citep{hayes20191000,RosFreixedes2022}.
% Finally, there is a subtle and important point about how recombination
% is represented: which genome in the organismal lifecycle
% do nodes represent?
% In principle, we can interpret nodes as representing genomes
% at any point in the life cycle [supplementary Figure ?]. Hudson
% views nodes as representing gametes~\citep{hudson1983properties}.
% However, we argue that the most intuitive
% approach is to let nodes represent the genomes present in diploid
% individuals in each generation, as in
% Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
% is not represented by a single ``recombination node'' (as in an eARG), but with
% two parent genomes between which recombination takes place
% (e.g.\ nodes \textsf{e} and \textsf{f}), and a
% single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
% edges that link this recombinant with its two parents carry the information
% concerning breakpoint location that would be associated with a
% recombination node in an eARG. [TODO finish up here with more
% clarity on what this means in terms of the required specifity
% of recombination events, and put in a forward ref to the simplification
% section where we talk about stacking up recombination events
% and their identifiability.]

The eARG and gARG definitions provided here are apparently very
similar, but, as we shall see, there are many advantages to decoupling
our representation of patterns of genetic inheritance
from the events that generated them.
For simplicity, throughout the rest of the paper we will contrast the
classical eARG with the gARG definition here.
However, it is important to note that many of the benefits of the
gARG encoding that we discuss are due to the encoding of inheritance via
edges rather than nodes.
Whether one regards nodes as genomes or events is essentially
a matter of interpretation, and unimportant for many purposes.

\section*{Inheritance and ancestral material}\label{Ancestry_resolution}
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
number of generations (ultimately from cell to cell). The edges are ``annotated''
with the genome coordinates over which inheritance occurs. There are
two ways in which we can interpret these annotations, and the
distinction is subtle and important.
The first interpretation is to regard the annotations as describing
the complete set of genomic intervals inherited by the descendant node from the
ancestor node. The second is to view the edge annotations as representing
the ``ancestral material''~\citep{wiuf1999ancestry,wiuf1999recombination}
of the sample that is propagated through
the graph from descendant to ancestor.
These alternatives are illustrated in Fig~\ref{fig-ancestry-resolution}.
